Local Type Inference for Polarised System F with Existentials

### 2024-04-12 oopsla24/tex/02-overview.tex
WIP: examples, overview

The standard \systemf types and terms must be polarized to be used in \fexists, 
we present the embedding formally in \cref{fig:polarization}.
For instance, the following types and terms of \systemf 
can be polarized in the given way:

\begin{figure}[h]
  \begin{align*}
    [[map]] &: [[↓∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺]] \\
    [[len]] &: [[↓∀α⁺.List α⁺ → ↑Int]] \\
    [[choose]] &: [[↓∀α⁺.α⁺ → α⁺ → ↑α⁺]] \\
    [[id]] &: [[↓∀α⁺.α⁺ → ↑α⁺]] \\
    [[auto]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
    [[autoP]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
  \end{align*}
  \caption{Polarization of standard \systemf terms}
  \label{fig:polarization-examples}
\end{figure}

Let us assume that $[[Φ]]$ is a context containing the terms from \cref{fig:polarization-examples}.
Then the following types are inferrable in \fexists:

\begin{align*}
  % &[[· ; Φ ⊢ [two, three, nine] : List Int]]\\
  &[[· ; Φ, x1:↓(↓iN → iN), x2:↓(↓iM → iM) ⊢ let x = choose (x1, x2); return x : ↑ ∃γ⁻.↓(↓γ⁻ → γ⁻)]]\\
  &[[· ; Φ ⊢ let x = choose (id, auto); return x : ↑ ∃γ⁻.↓γ⁻]]\\
  &[[· ; Φ ⊢ let x = choose (id, autoP); return x : ↑ ∃γ⁻.↓γ⁻]]\\
\end{align*}

The focus of our system is the \emph{local} type inference \cite{pierce2000:local},
which has certain limitations.
In particular, if an argument of a function is polymorphic,
its polymorphic variables are not instantiated by the inference algorithm. 
For example, one would expect the following to be inferrable:
$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$.
However, to infer this, the $[[id]]$ must be instantiated to $[[Int → ↑Int]]$, 
which requires the information to be propagated from the neighboring branch 
of the syntax tree ($[[ [two, three, nine] ]]$), \ie bypass the locality.

However, the polymorphic arguments of the function are instantiated, 
and thus, the annotated version of the same term infers the type successfully:
$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$.

@@

\begin{figure}[h]
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottiN\ottinterrule
    }\\
    \ottgrammartabular{
      \ottiP\ottinterrule
    }
    \columnbreak
  \end{multicols}
  \caption{Declarative Types of \fexists}
  \label{fig:declarative-types}
\end{figure}

@@

\begin{figure}[h]
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottc\ottinterrule
    }

    \ottgrammartabular{
      \ottv\ottinterrule
    }
  \end{multicols}
  \caption{Declarative Terms of \fexists}
  \label{fig:declarative-terms}
\end{figure}

@@

>>>>

In general, any term and type of \systemf can be embedded into \fexists in
several ways, because at each moment one can choose the positive or negative
polarity of the subterm. The consistent term-level and type-level embeddings
(that preserve typing) will be covered in detail in
\cref{sec:rel-to-systemf}. Now, we discuss several examples of standard
\systemf terms and types and their polarization in \fexists.

\begin{figure}[h]
  \begin{align*}
    [[map]] &: [[↓(∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺)]] \\
    [[len]] &: [[↓(∀α⁺.List α⁺ → ↑Int)]] \\
    [[choose]] &: [[↓(∀α⁺.α⁺ → α⁺ → ↑α⁺)]] \\
    [[id]] &: [[↓(∀α⁺.α⁺ → ↑α⁺)]] \\
    [[auto]] &: [[↓(↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺))]]
  \end{align*}
  \caption{Polarization of standard \systemf terms}
  \label{fig:polarization-examples}
\end{figure}

Let us assume that $[[Φ]]$ is a context containing the terms from \cref{fig:polarization-examples}.
Then the following types are inferrable in \fexists.

\paragraph{Application of a Polymorphic Function}
  A polymorphic function can be applied to a value of a concrete type.
  In this case, the quantified variables are instantiated to make the application type-check.
  For instance, the polymorphic $[[id]]$ can be applied to any positive type, 
  and the resulting type will be inferred by the algorithm:
  $$[[· ; Φ ⊢ let x = id(fortyTwo); return x : ↑Int]]$$

\paragraph{Polymorphic Self-application}
  Unlike predicative systems, \fexists
  permits self-application of polymorphic functions.
  For instance, $[[id]]$ can be applied to itself
  resulting in a polymorphic type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

\paragraph{Non-example: Polymorphic Arguments}
The focus of our system is the \emph{local} type inference \cite{pierce2000:local},
which has certain limitations.
In particular, if an argument of a function is polymorphic,
its polymorphic variables are \emph{not} instantiated by the inference algorithm. 
For example, one would expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch 
of the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

The annotated version of the same term infers the type successfully, 
since all the polymorphic variables of the arguments are instantiated:
$$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$$

\paragraph{Inferring the Common Supertype}
  The function $[[choose]]$ takes two arguments of the same
  polymorphic type. To apply it to two values of different (concrete) types, 
  the inference algorithm must find the minimal type they can be coerced 
  to---the least common supertype.

  Let us assume that in $[[Φ]]$, we have two identity-like functions 
  \mbox{$[[id_iM]] : [[↓(↓iM → iM)]]$} and \mbox{$[[id_iN]] : [[↓(↓iN → iN)]]$}.
  Then their least common supertype---the existential $[[∃γ⁻.↓(↓γ⁻ → γ⁻)]]$
  is inferrable:
  $$[[· ; Φ ⊢ let x = choose (id_iN, id_iM); return x : ↑ ∃γ⁻.↓(↓γ⁻ → γ⁻)]]$$

\paragraph{Inferring a general type}
  As mentioned before, the local inference does not instantiate the polymorphic
  types in the argument position.
  Because of that, the application 
  \mbox{$[[let x = choose (id, auto); return x ]]$} cannot infer 
  $[[↑↓(↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺))]]$.
  However, the inference succeeds and infers a less informative type:
  $$[[· ; Φ ⊢ let x = choose (id, auto); return x : ↑ ∃γ⁻.↓γ⁻]]$$

@@

\begin{figure}[h]
  \begin{minipage}[t]{0.6\textwidth}
    \begin{supertabular}{ll p{0.8\textwidth}}
      \ottiNInline\\
    \end{supertabular}
  \end{minipage}
  \begin{minipage}[t]{0.35\textwidth}
    \begin{supertabular}{ll p{0.8\textwidth}}
      \ottiPInline\\
    \end{supertabular}
  \end{minipage}

  \caption{Declarative Types of \fexists}
  \label{fig:declarative-types}
\end{figure}

@@

\begin{figure}[h]
  \begin{supertabular}{ll p{0.8\textwidth}}
    \ottcInline\\
    \ottvInline\\
  \end{supertabular}
  \caption{Declarative Terms of \fexists}
  \label{fig:declarative-terms}
\end{figure}

@@

The inference algorithm for \fexists is \emph{local}: it has a limited scope of
the inference information and does not propagate it between far-apart branches
of the syntax tree. Nevertheless, many difficulties appear in the inference of
polymorphic and existential types. Next, we discuss the most challenging of
these difficulties and how they are addressed in the algorithm.

\paragraph{Subtyping} Most of the difficulty of the inference algorithm is
concentrated in the subtyping relation. This is the part of the system where
most of the polymorphic instantiation happens. In particular, it is the
subtyping relation that allows us to use the polymorphic terms in the context
where their concrete instantiations are expected.

To check whether one type is a subtype of another, the algorithm introduces
\emph{algorithmic} type variables ($[[α̂±]], [[β̂±]], \dots$)--- the
`placeholders' representing the polymorphic variables whose instantiation is
postponed. This way, the problem of $[[∀α⁺.↑α⁺ ≤ ↑Int]]$ becomes the problem
of finding $[[α̂⁺]]$ such that $[[↑α̂⁺ ≤ ↑Int]]$.

\newcommand{\defiff}{%
  \mathrel{\vbox{\offinterlineskip\ialign{%
    \hfil##\hfil\cr
    $\scriptscriptstyle\text{def}$\cr
    %\noalign{\kern0ex}
    $\iff$\cr
}}}}

\paragraph{Equivalence}
Our system introduces another complexity: the equivalence induced by subtyping
($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
straightforward. This relation extends beyond just alpha-equivalent types. As
the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
the equivalence relation must permit the permutations of the quantifiers. For
example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
equivalence permits the removal/addition of the unused quantifiers: $[[· ⊢
∀α⁺.↑Int ≈ ↑Int]]$.

In order to manage the complexity of non-trivial equivalence, we implement
a type normalization algorithm. It effectively reduces the mutual subtyping to
a simpler alpha-equivalence. The normalization works by recursively
eliminating unused quantifiers and reordering the remaining ones into a
canonical permutation.

\paragraph{Least Upper Bound}
Eventually, the subtyping algorithm reduces the initial subtyping problem to a set of
constraints that need to be resolved. At this moment, another problem arises:
as one variable can occur in multiple constraints, the resolution becomes non-trivial.
In particular, to resolve the conjunction of $[[α̂⁺ ≥ iP]]$ and $[[α̂⁺ ≥ iQ]]$,
we would need to find the least upper bound of $[[iP]]$ and $[[iQ]]$.

In many type systems, the least upper bound boils down to the trivial syntactic equality.
Because of that, the set of constraints is typically resolved using \emph{unification}.
However, this problem is trickier in the presence of existential types. For example, 
in \fexists, the least upper bound of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[∃α⁻.↓α⁻]]$,
because the quantified $[[α⁻]]$ can be instantiated to either $[[↑Int]]$ or $[[↑Bool]]$.

To find this least upper bound, we use \emph{anti-unification}---the dual of
unification. While unification finds \emph{the most general instance} of two
given patterns, the anti-unification finds \emph{the most detailed} (under the
restrictions of the system) \emph{pattern} that matches two given types. In the
example above, the anti-unifier of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[↓α̂⁻]]$,
and in \fexists, this unifier is the most specific (in particular because
$[[↓↑α̂⁺]]$ is not considered as only \emph{negative} variables are
existentially quantified).

\paragraph{Impredicativity}
Another difficulty that hinders the constraint resolution arises from the
impredicativity of the system---the ability to quantify over the types that are 
themselves polymorphic. In the impredicative polymorphic system (even
with only $[[∀]]$-quantifiers), the naturally formulated subtyping is
undecidable, which follows from the undecidability of second-order unification.
All the more the undecidability is evident in the presence of existential types
and subtyping constraints. For instance, in \fexists the constraint 
$[[α̂⁺ :≥ ↓↑α̂⁺]]$ would have a surprising impredicative solution $[[α̂⁺]] = [[∃α⁻.↓α⁻]]$.

To maintain the decidability in the presence of impredicativity, we carefully
restrict the system using the polarity stratification. One of the main
constraints that we put on subtyping is the invariance of the shift operators:
the subtyping $[[↑iP ≤ ↑iQ]]$ is only allowed if $[[iQ ≥ iP]]$
\emph{and} $[[iP ≥ iQ]]$. These constraints do not trivialize the system: we
still have non-trivial upper bounds, and thus, need to use anti-unification to
find them. However, it prevents such examples as $[[α̂⁺ :≥ ↓↑α̂⁺]]$, and in
fact, makes the constraint resolution decidable.

### 2024-04-12 oopsla24/tex/03-declarative-system.tex
WIP: examples, overview

\paragraph{Invariant Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants make the system equivalent to \systemf, 
and thus, undecidable. 
However, after certain changes 
\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form,
thereby increasing the expressiveness of the system. These
changes are discussed in \cref{sec:weakening-invariant}.

>>>>

\paragraph{Invariant Shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that the subtyping on shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}.
Relaxing both of these invariants make the system equivalent to \systemf, and
thus, undecidable. However, after certain changes
\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form, thereby
increasing the expressiveness of the system. These changes are discussed in
\cref{sec:weakening-invariant}.

### 2024-04-12 oopsla24/tex/04-algorithm.tex
WIP: examples, overview

\begin{figure}[h]
  \begin{multicols}{2}
      Negative Algorithmic Variables\\
      $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots\\
      \columnbreak
      Positive Algorithmic Variables\\
      $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots\\
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
      \ottuNShort
      \columnbreak
      \ottuPShort
  \end{multicols}
  \hfill\\
  Algorithmic Type Context\\
   $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \hfill\\
  Constraint Type Context\\
   $[[Θ]] ::= \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \caption{Algorithmic Syntax}
  \label{fig:algo-syntax}
\end{figure}

@@

  \begin{figure}[h]
    \begin{multicols}{2}
      \ottgrammartabular { 
      \ottscE 
      \ottinterrule
      \ottruleheadOneLine
        {[[SC]]}{::=} {\ottcom{Constraint Set}}
        {\{[[scE1]], \dots, [[scEn]]\}}\ottprodnewline
      }

@@

      \ottgrammartabular { 
      \ottucE 
      \ottprodnewline
      \ottinterrule
      \ottruleheadOneLine
        {[[UC]]}{::=} {\ottcom{Unification Constraint Set}}
        {\{[[ucE1]], \dots, [[ucEn]]\}}\ottprodnewline
      }
    \end{multicols}

>>>>

\begin{figure}[h]
  \begin{multicols}{2}
      Negative Algorithmic Variables\\
      $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots\\
      \columnbreak
      Positive Algorithmic Variables\\
      $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots\\
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
      \ottuNShort
      \columnbreak
      \ottuPShort
  \end{multicols}
  \hfill\\
  Algorithmic Type Context\\
   $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \hfill\\
  Constraint Type Context\\
   $[[Θ]] \Coloneqq \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \caption{Algorithmic Syntax}
  \label{fig:algo-syntax}
\end{figure}

@@

  \begin{figure}[h]
    \begin{multicols}{2}
      \ottgrammartabular { 
      \ottscE 
      \ottinterrule
      \ottruleheadOneLine
        {[[SC]]}{\Coloneqq} {\ottcom{Constraint Set}}
        {\{[[scE1]], \dots, [[scEn]]\}}\ottprodnewline
      }

@@

      \ottgrammartabular { 
      \ottucE 
      \ottprodnewline
      \ottinterrule
      \ottruleheadOneLine
        {[[UC]]}{\Coloneqq} {\ottcom{Unification Constraint Set}}
        {\{[[ucE1]], \dots, [[ucEn]]\}}\ottprodnewline
      }
    \end{multicols}
